---
title: "Phyloseq"
author: "Janelle"
date: "2024-04-18"
output: pdf_document
---
# to intsall phyloseq packages
```{r}
source('http://bioconductor.org/biocLite.R')
biocLite('phyloseq')
```

```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("phyloseq")
```

# load required packages
```{r}
library(phyloseq)
library(dplyr) # alternative installation of the %>%
library(BiMiCo)
library(ggplot2)
library(devtools)
library(vegan)
library(MicEco)
```

# Load taxa and seqtab.nochim
```{r}
load("RData/taxa.RData")
load("RData/seqtab.nochim,RData")
```

# import metadata
```{r}
metadata<-read.csv("sample-metadata.csv", header=TRUE, row.names=1)
```

# create phyloseq object
```{r}
#make sure the seqtab.nochim and taxa objects are loaded
physeq <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows=FALSE),
                   sample_data(metadata),
                   tax_table(taxa))
physeq
```

# transform sample counts
```{r}
# converts from raw to abundance
physeq <- transform_sample_counts(physeq, function(abund) 1*(abund>0))
physeq
```
# Remove the sequence itself and replace with ASV
```{r}
dna <- Biostrings::DNAStringSet(taxa_names(physeq))
names(dna) <- taxa_names(physeq)
physeq <- merge_phyloseq(physeq,dna)
taxa_names(physeq) <- paste0("ASV", seq(ntaxa(physeq)))
physeq
```

# remove mitochondria and chloroplast matches. remove all non bacterial sequences
```{r}
physeq <- physeq %>% subset_taxa(Family!= "Mitochondria" | is.na(Family) & Order!="Chloroplast" | is.na(Order))
physeq
```

# save physeq objects to load later
```{r}
save(physeq, file="RData/physeq.RData")
```

# load physeq objects start here
```{r}
load("RData/physeq.RData")
```

# plot bar graph based on phylum
```{r}
plot_bar(physeq, fill = "Phylum") + geom_bar(aes(color=Phylum, fill=Phylum), stat="identity", position="fill")
```

# in console
library(ggplot2)

#install package ggplot2 in console (add to top w/ other libraries)
install.packages("ggplot2")

# create a barplot of relative abundance
```{r}
#convert to relative abundance
physeq_relabund <- transform_sample_counts(physeq, function(x) x / sum(x))

# barplot
plot_bar(physeq_relabund, fill = "Phylum") + geom_bar(aes(color=Phylum, fill=Phylum), stat="identity", position="fill") + facet_wrap(~body.site, scales = "free")
```
# Bar plots of Abundance per individual samples by body sit (Phylum-Merge <5%)

# plot alpha diversity based on body.site
```{r}
plot_richness(physeq, x="body.site", color= "subject", measures=c("Observed","Simpson", "Shannon"))
```

# plot alpha diversity based on reported.antibiotic.usage
```{r}
plot_richness(physeq, x="reported.antibiotic.usage", color= "body.site", measures=c("Observed","Simpson", "Shannon"))
```

# plot alpha diversity based on subject 
```{r}
plot_richness(physeq, x="subject", color= "body.site", measures=c("Observed","Simpson", "Shannon"))
```

# test for normality - pulls out sample and calculates its richness
```{r}
alpha <- estimate_richness(physeq, measures=c("Observed","Simpson", "Shannon"))

# shapiro-wilk
observed <- shapiro.test(alpha$Observed)
shannon <- shapiro.test(alpha$Shannon)
simpson <- shapiro.test(alpha$Simpson)

#print
print(observed)
print(shannon)
print(simpson)
```

# create data frames for statistical analyses 
```{r}
# extract sample information from the physeq object; pulls out all the sample information (i.e. names)
samples <- sample_data(physeq)

# If samples is a phyloseq sample_data object, convert it to a data frame
if (class(samples) == "sample_data") {
samples <- data.frame(sample_data(samples))
}

# add a column to alpha with sample names
alpha$sample <- rownames(alpha)

# merge alpha diversity data and sample data
alpha <- merge(alpha, samples, by = "sample")
```

# perform stats based on subject
```{r}
# perform t/wilcox tests for each biodiversity index
test_observed <- wilcox.test(Observed ~ subject, data = alpha)
test_simpson <- wilcox.test(Simpson ~ subject, data = alpha)
test_shannon <- wilcox.test(Shannon ~ subject, data = alpha)

# printing the results
print(test_observed)
print(test_simpson)
print(test_shannon)
```
# perform stats based on reported.antibiotic.usage
```{r}
# perform t/wilcox tests for each biodiversity index
test_observed <- wilcox.test(Observed ~ reported.antibiotic.usage, data = alpha)
test_simpson <- wilcox.test(Simpson ~ reported.antibiotic.usage, data = alpha)
test_shannon <- wilcox.test(Shannon ~ reported.antibiotic.usage, data = alpha)

# printing the results
print(test_observed)
print(test_simpson)
print(test_shannon)
```

# perform stats based on body.site
```{r}
kruskal.test(Simpson ~ body.site, data=alpha)
pairwise.wilcox.test(alpha$Simpson, alpha$body.site, p.adjust.method="holm")
kruskal.test(Observed ~ body.site, data=alpha)
pairwise.wilcox.test(alpha$Observed, alpha$body.site, p.adjust.method="holm")
shannonanova <- aov(Shannon ~ body.site, data=alpha)
summary(shannonanova)
TukeyHSD(shannonanova)
```

# Remove taxa with relative abundance <0.005%
```{r}
minTotRelAbun = .00005
x = taxa_sums(physeq)
keepTaxa = (x / sum(x)) > minTotRelAbun
physeqprune = prune_taxa(keepTaxa, physeq)
physeqprune
```
# Number of shared ASVs body.site (found in 25% or more)
```{r}
bodysite=ps_venn(
  physeqprune,
  "body.site",
  fraction = 0.25,  
  weight = FALSE,
  relative = TRUE,
  plot = TRUE
)
bodysite
```
# Bray Curtis Calculation
```{r message=FALSE}
set.seed(777) # calculate differences, to make calculations more 
dist = phyloseq::distance(physeqprune, method="bray", weighted=TRUE)
ordination = ordinate(physeqprune, method="NMDS", distance=dist)
```

# Bray Curtis Subject Plot
```{r}
braysubject=plot_ordination(physeqprune, ordination, color="subject") +
theme_classic() +
theme(strip.background = element_blank()) + stat_ellipse(aes(group=subject))
braysubject
```
# Bray Curtis Body Site Plot
```{r}
braybodysite=plot_ordination(physeqprune, ordination, color="body.site") +
theme_classic() +
theme(strip.background = element_blank()) + stat_ellipse(aes(group=body.site))
braybodysite
```

```{r}
brayreportedantibioticusage=plot_ordination(physeqprune, ordination, color="reported.antibiotic.usage") +
theme_classic() +
theme(strip.background = element_blank()) + stat_ellipse(aes(group=reported.antibiotic.usage))
brayreportedantibioticusage
```
# Bray Curtis Subject Statistics 
```{r}
adonis2(dist ~ sample_data(physeqprune)$subject)
```
# Bray Curtis Body Site Statistics 
```{r}
adonis2(dist ~ sample_data(physeqprune)$body.site)
```
# Bray CurtisReported Antibiotic Resistance Statistics 
```{r}
adonis2(dist ~ sample_data(physeqprune)$reported.antibiotic.usage)
```




